Updates a Procrustes and a PCA

### 2025-04-28 Analysis/17-update_procrustes_and_PCA.Rmd
Update 17-update_procrustes_and_PCA.Rmd

@@

## The following are missing:
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_africana_14_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_africana_19_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_coelebs_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_21_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_5_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_moreletti_11_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_palmae_13_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_teydea_polatzeki_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_teydea_teydea_10_intraspecific/: empty"

## Converting into an array
individuals_array <- simplify2array(all_individuals)

## Project them! (*-1 is to flip it around)
updated_pca <- update.gpa.pca(whole_gpa, whole_gpa$rotated[,,c(2819, 2820)], scale.pca = FALSE)

@@

```{r}
library(Morpho)

## Creating some new specimens (two old ones, for benchmarking the method)
new_spec <- array(c(original_landmarks[,,1],
                    original_landmarks[,,2]), dim = c(10, 3, 2))

## Project the new specimens on the original GPA
new_gpa <- Morpho::align2procSym(proc_orig, new_spec, orp = TRUE)

## Slightly off?
plot.offset <- function(id, new_gpa, old_gpa, main) {
    main <- paste0(main, "\nTotal offset = ", sum(sqrt((new_gpa[, 1, id] - old_gpa[, 1, id])^2 + (new_gpa[, 2, id] - old_gpa[, 2, id])^2)))
    plot(new_gpa[, c(1,2), id], main = main, xlab = "", ylab = "")
    points(old_gpa[, c(1,2), id], pch = 19, cex = 0.5)
}

op <- par(mfrow = c(2,1))
plot.offset(id = 1, new_gpa, proc_orig$rotated, main = "Spec 1 projected onto itself")
plot.offset(id = 2, new_gpa, proc_orig$rotated, main = "Spec 2 projected onto itself")
par(op)

data(boneData)
# run procSym on entire data set
proc <- procSym(boneLM)
# this is the training data
array1 <- boneLM[,,1:60]
proc1 <- procSym(array1)
newdata <- boneLM[,,1:2]
newalign <- align2procSym(x = proc1, newdata = newdata)

plot.offset(id = 1, newalign, proc1$rotated, main = "Spec 1 projected onto itself")

TODO: try doing the alignement of some species on the edges of the traitspace and see if it falls back on the edges or if it goes to the middle again.

TODO: check with the `morphospace` package?

>>>>

Checking all over again with `princomp`

```{r}
## New test
library(Morpho)
library(landvR) # install from devtools::install_github("TGuillerme/landvR")

## New species (three to anchor and two new ones)
new_spec <- array(data = c(original_landmarks[,,1],
                           original_landmarks[,,2],
                           original_landmarks[,,3],
                           apply(original_landmarks, c(1,2), median),
                           apply(original_landmarks, c(1,2), min)),  dim = c(10, 3, 5))
dimnames(new_spec)[[3]] <- c(paste0("old_", c("spec1", "spec2", "spec3")), "median", "min")

## Run the PCA for the original data
pca_orig <- prcomp(landvR::array.to(proc_orig$rotated, to = "matrix"))
princomp_orig <- princomp(landvR::array.to(proc_orig$rotated, to = "matrix"))

## THIS PIPELINE SEEMS TO WORK:

## Test the reprojection using princomp
new_gpa <- Morpho::align2procSym(proc_orig, new_spec)
new_points <- stats::predict(princomp_orig, newdata = array.to(new_gpa, to = "matrix"))
plot(princomp_orig$scores[,c(1,2)], col = "grey")
points(new_points[,c(1,2)], col = "orange", pch = 19)
points(princomp_orig$scores[c(1,2,3),c(1,2)], col = "blue", pch = 21)

## Thest using prcomp
new_gpa <- Morpho::align2procSym(proc_orig, new_spec)
new_points <- stats::predict(pca_orig, newdata = array.to(new_gpa, to = "matrix"))
plot(pca_orig$x[,c(1,2)], col = "grey")
points(new_points[,c(1,2)], col = "orange", pch = 19)
points(pca_orig$x[c(1,2,3),c(1,2)], col = "blue", pch = 21)

## Test with pipeline
updated_pca <- update.gpa.pca(proc_orig, new_spec)*-1 # TODO: fix the -1
plot(proc_orig$PCscores[,c(1,2)])
points(updated_pca[,c(1,2)], col = "grey", pch = 19)
points(updated_pca[c(81:85),c(1,2)], col = "orange", pch = 19)
points(updated_pca[c(81:83),c(1,2)], col = "blue", pch = 21)

## Not really equal though!
expect_equal(updated_pca[-c(81:85),c(1,2)], proc_orig$PCscores[,c(1,2)])

@@

## Converting into an array
new_individuals <- simplify2array(all_individuals)

## Project them! (*-1 is to flip it around)
updated_pca <- update.gpa.pca(whole_gpa, new_individuals)*-1

# No pipeline
# Get the CPA
original_data <- whole_gpa
new_landmarks <- new_individuals
original_pca <- prcomp(landvR::array.to(original_data$rotated, to = "matrix")) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
# original_pca <- prcompfast(original_data$tan) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
## Project the new specimens on the original GPA
new_data <- Morpho::align2procSym(original_data, new_landmarks)
## Project the new GPAed specimens onto the PCA
new_pca <- stats::predict(original_pca, newdata = array.to(new_data, to = "matrix"))
## Combine the new and old matrices for the output PCA
updated_pca <- rbind(original_pca$x, new_pca) *-1

@@
